Leaf mass per area and plant size drive differences in above-ground biomass distribution among woody plant functional types

### 2016-02-04 manuscript.Rnw
caption fix
More edits (intro) and references added

%The Summary for research papers, which must be usable as a stand-alone document, must not exceed 200 words and should be organized using four bullet points to indicate (1) the research conducted, including the rationale, (2) methods, (3) key results and (4) the main conclusion, including the key points of discussion. 
\begin{itemize}
\item The partitioning of biomass into leaves and stems is an important determinant of the residence time of carbon in vegetation. We tested the hypothesis that biomass distribution in leaves and stems is related to the leaf mass per area, an important leaf trait.
\item We used a large compilation of plant biomass and size observations, including observations of \Sexpr{nrow(baad_all$data)} individuals on \Sexpr{length(unique(baad_all$data$speciesMatched))} species. We used a combination of semi-parametric methods and variance partitioning to test for influence of PFT, plant height, and climate on biomass distribution and three component variables.
\item The ratio of leaf mass to woody above-ground mass ($M_F/A_S$) varied strongly between PFTs (as defined by deciduous vs. evergreen and gymnosperm vs. angiosperm). We found that $M_F/A_S$, once corrected for plant height, was proportional to leaf mass per area across PFTs. As a result, the PFTs did not differ in the amount of leaf area supported per unit above ground biomass. Climate consistently explained very little additional variation, demonstrating very weak control of climate on biomass distribution, at a given plant size. 
\item Combined, these results demonstrate consistent patterns in above-ground biomass distribution and leaf area relationships among major woody plant functional types, and can be used to further constrain global vegetation models.
\end{itemize}

@@

The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation \citep{ise2010, dekauwe2014,friend2014}. Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \citep{ryanlaw2005}. Globally, forests store approximately 360Pg of carbon in living biomass \citep{panetal2011}, equivalent to almost 40 years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}. Reducing uncertainties about biomass partitioning and carbon residence times is a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014,negron2015}.

One puzzling aspect of the large inter-specific variation in biomass distribution between leaves and stems is the much larger leaf mass in gymnosperms compared to angiosperms (at a given stem mass) \citep{oneilldeangelis1981,enquistniklas2002,reich2014,poorter2012, poorter2015}. One possible explanation is that gymnosperms compensate for their low leaf mass per area (LMA) \citep{poorter2009} by maintaining a large amount of leaf biomass. If this is the case, total leaf area per unit above-ground biomass would be less variable between angiosperms and gymnosperms than leaf mass. Some oft-cited studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}, but recent large compilations on biomass distribution did not include measurements of leaf area in addition to mass to test this hypothesis \citep{poorter2012,poorter2015,reich2014}. The idea that high total leaf mass may compensate for low LMA would also explain the difference between evergreen (high LMA, high leaf mass) and deciduous (low LMA, low leaf mass) species \citep{poorter2015}. Here we set out to directly test the hypothesis that across PFTs, $M_F/M_S$ is related to LMA with a global database that includes observations of leaf area as well as leaf mass measured on the same plants.

@@

We excluded glasshouse and common garden studies, including only field-grown woody plants (including natural vegetation, unmanaged and managed plantations). Excluding plantations from the analysis did not affect the main result, but far fewer data were available for evergreen gymnosperms. We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms. We excluded  deciduous gymnosperms because few data were available. To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature (MAT) and precipitation (MAP) from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}). MAT and MAP were also extracted for all study locations. Potential evapotranspiration (PET) was extracted from the Global-PET database (http://www.cgiar-csi.org, \citet{zomer2008}), and an aridity index was calculated as PET/MAP.

@@

\begin{equation}
D_S = D_{Sbh} H^a / (H - H_{bh})^a
\end{equation}

where $D_S$ is the basal stem diameter (m), $D_{Sbh}$ stem diameter at breast height, $H_{bh}$ the height at which $D_{Sbh}$ was measured (typically 1.3 or 1.34m), and $a$ was further expressed as a function of plant height:

\begin{equation}
a = c_0H^{c_1}
\end{equation}

@@

Because, similar to \citep{poorter2015}, none of the studied relationships were linear on a log-log scale, we did not use standard allometric equations to study relationships between size and biomass variables. Instead, we used generalised additive models (GAM), a semi-parametric modelling approach that makes no assumption about the shape of the relationships. GAMs were used to visualize the relationships between biomass and plant size variables, and to estimate variables and their confidence intervals such as $M_F/M_S$ at a common plant height. In all fitted GAMs, we used a cubic regression spline. For the smoothed term in the model (plant height), we used up to 3 or 4 degrees of freedom, which resulted in biologically realistic smoothed relationships. Within the GAM, we used a penalised regression smoother \citep{wood2006} to allow the final degree of smoothness to be estimated from the data. In all fitted GAMs, we used species-dataset combination as a random effect. All variables (except MAP and MAT) were log-transformed prior to analysis.

@@

Previous interpretations of differences in $M_F/M_S$ between gymnosperms and angiosperms have focussed on differences in leaf life-span (LL) \citep{enquistniklas2002,poorter2012}. Indeed, if all else equal, longer LL in gymnosperms would result in higher total leaf mass. It is however difficult to separate cause from effect in the relationships between LL, LMA, and $M_F/M_S$, particularly because LL and LMA are correlated across species and PFTs in a nearly proportional way \citet{wright2004}.

@@

We stress that although our results show some convergence between PFTs in leaf area relationships, we do not dismiss the very large residual variation that is yet to be explained and understood (Fig. 6). Our purpose was to test hypotheses related to differences between PFTs, plant height, and climate, yet clearly more work can be done to understand the variation between species and sites witin groups. Some of the residual variation probably has to do with site fertility (CITES), or local differences in water availability, perhaps mediated by seasonality, that were not picked up by our coarse climate variables.

@@

\clearpage
\section{Figures}

@@

\begin{figure}[h!]
    \centering
    \ifincludefigs 
    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
    \fi
    \caption{Dominant woody PFTs differ in above-ground biomass distribution due to underlying differences in leaf mass per area. (a) Leaf mass per unit woody above-ground biomass ($M_F/M_S$) by PFT. Each symbol is an individual plant. Lines are generalised additive model fits. (b) and (c) $M_F/M_S$ and leaf area ratio ($A_F/M_S$) at the average plant height in the dataset, estimated from panel (a). Error bars are 95\% confidence intervals. Letters denote significant differences (at $\alpha = 0.05$).}
    \label{fig:figlmf}
\end{figure}

>>>>

%The Summary for research papers, which must be usable as a stand-alone document, must not exceed 200 words and should be organized using four bullet points to indicate (1) the research conducted, including the rationale, (2) methods, (3) key results and (4) the main conclusion, including the key points of discussion. 
\begin{itemize}
\item We tested the hypothesis that biomass distribution in leaves and stems is related to the leaf mass per area, an important leaf trait, across major woody plant functional types (PFTs) (as defined by deciduous vs. evergreen and gymnosperm vs. angiosperm).  
\item We used a large compilation of plant biomass and size observations, including observations of \Sexpr{nrow(baad_all$data)} individuals on \Sexpr{length(unique(baad_all$data$speciesMatched))} species. We used a combination of semi-parametric methods and variance partitioning to test for influence of PFT, plant height, and climate on biomass distribution and three component variables.
\item The ratio of leaf mass to woody above-ground mass ($M_F/M_S$) varied strongly between PFTs. We found that $M_F/M_S$, once corrected for plant height, was proportional to leaf mass per area across PFTs. As a result, the PFTs did not differ in the amount of leaf area supported per unit above ground biomass. Climate consistently explained very little additional variation in biomass distribution at a given plant size. 
\item Combined, these results demonstrate consistent patterns in above-ground biomass distribution and leaf area relationships among major woody plant functional types, and can be used to further constrain global vegetation models.
\end{itemize}

@@

The distribution of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation \citep{ise2010, dekauwe2014,friend2014}. Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \citep{ryanlaw2005}. Globally, forests store approximately 360Pg of carbon in living biomass \citep{panetal2011}, equivalent to almost 40 years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}. Reducing uncertainties about biomass distributionand carbon residence times is a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014,negron2015}, in particular because projections by global vegetation models (GVMs) are particularly sensitive to this component \citep{friend2014,ise2010}.

Perhaps the biggest challenge is to capture the combined responses to environmental change of the more than 250,000 plant species comprising the world's vegetation. While most plants have the same basic resource requirements and physiological function, large differences exist among species in the amount of biomass invested in different tissues (leaves, stems, roots). To attempt to understand the source of this functional diversity, it is customary to consider only a few archetypal plant functional types (PFTs) \citep{harrisonetal2010,wullschleger2014}, assumed to differ in key attributes. Previous work has shown consistently large differences in biomass distribution and related traits among major PFTs \citep{poorter2009,enquistniklas2002,poorter2012,poorter2015,reich2014,mccarthyetal2007}, but these differences are yet to be fully understood.

One puzzling aspect of the large variation in biomass distribution among PFTs is the much larger leaf mass in gymnosperms compared to angiosperms (at a given stem mass) \citep{oneilldeangelis1981,enquistniklas2002,reich2014,poorter2012, poorter2015}. A possible explanation is that gymnosperms compensate for their low leaf mass per area (LMA) \citep{poorter2009} by maintaining a large amount of leaf biomass, to maintain sufficient total leaf area. If this is the case, total leaf area per unit above-ground biomass would be less variable between angiosperms and gymnosperms than leaf mass. Some oft-cited studies have however assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}, but recent large compilations on biomass distribution did not include measurements of leaf area in addition to mass to test this hypothesis \citep{poorter2012,poorter2015,reich2014}. The idea that high total leaf mass may compensate for low LMA would also explain the difference between evergreen (high LMA, high leaf mass) and deciduous (low LMA, low leaf mass) species \citep{poorter2015}. Here we set out to directly test the hypothesis that across PFTs, $M_F/M_S$ is related to LMA with a global database that includes observations of leaf area as well as leaf mass measured on the same plants.

@@

We included only field-grown woody plants in our analysis (including natural vegetation, unmanaged and managed plantations), excluding glasshouse and common garden studies. Excluding plantations from the analysis did not affect the main result, but far fewer data were available for evergreen gymnosperms. We considered three PFTs : evergreen angiosperms, evergreen gymnosperms, and deciduous angiosperms. We excluded  deciduous gymnosperms because few data were available. To assess the coverage of the global climate space by the dataset, we extracted mean annual temperature (MAT) and precipitation (MAP) from Worldclim \citep{worldclim}, excluding areas without woody vegetation (taken from the global land cover database GLC-SHARE \citep{lathamglobal}). MAT and MAP were also extracted for all study locations. Potential evapotranspiration (PET) was extracted from the Global-PET database (http://www.cgiar-csi.org, \citet{zomer2008}), and an aridity index was calculated as PET/MAP.

@@

\begin{equation}
D_S = D_{Sbh} H^c / (H - H_{bh})^c
\end{equation}

where $D_S$ is the basal stem diameter (m), $D_{Sbh}$ stem diameter at breast height, $H_{bh}$ the height at which $D_{Sbh}$ was measured (typically 1.3 or 1.34m), and \emph{c} was further expressed as a function of plant height:

\begin{equation}
c = c_0H^{c_1}
\end{equation}

@@

Because, similar to \citep{poorter2015}, none of the studied relationships were linear on a log-log scale, we did not use standard allometric equations to study relationships between size and biomass variables. Instead, we used generalised additive models (GAM), a semi-parametric modelling approach that makes no assumption about the shape of the relationships. GAMs were used to visualize the relationships between biomass and plant size variables, and to estimate variables and their confidence intervals such as $M_F/M_S$ at a common plant height. In all fitted GAMs, we used a cubic regression spline. For the smoothed term in the model (plant height), we used up to 3 or 4 degrees of freedom, which resulted in biologically realistic smoothed relationships. Within the GAM, we used a penalised regression smoother \citep{wood2006} to allow the final degree of smoothness to be estimated from the data. In all fitted GAMs, we used species-dataset combination as a random effect. To homogenize variance, all variables (except MAP and MAT) were log-transformed prior to analysis.

@@

Previous interpretations of differences in $M_F/M_S$ between gymnosperms and angiosperms have focussed on differences in leaf life-span (LL) \citep{enquistniklas2002,poorter2012}. Indeed, if all else equal, longer LL in gymnosperms would result in higher total leaf mass. Further supporting this explanation, \citet{reich1992} showed a positive correlation between LL and total stand leaf biomass across diverse forest stands. It is however difficult to separate cause from effect in the relationships between LL, LMA, and $M_F/M_S$, particularly because LL and LMA are correlated across species and PFTs in a nearly proportional way \citet{wright2004}.

@@

We stress that although our results show some convergence between PFTs in leaf area relationships, we do not dismiss the very large residual variation that is yet to be explained (Fig. 6). Our purpose was to test hypotheses related to first-order differences between PFTs, plant height, and climate, yet clearly more work can be done to understand the variation between species and sites witin groups. Some of the residual variation is likely explained by site fertility \citet{vanninen2005}, differences in stand density \citet{ilomaki2003}, social status within stands \citet{makela2006}, or local differences in water availability (possibly mediated by seasonality at a given MAP). Future work will focus on smaller-scale explanatory variables such as the ones mentioned here.

@@

\clearpage
\section*{Figures}

@@

\begin{figure}[h!]
    \centering
    \ifincludefigs 
    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
    \fi
    \caption{Dominant woody PFTs differ in above-ground biomass distribution due to underlying differences in leaf mass per area. (a) Leaf mass per unit woody above-ground biomass ($M_F/M_S$) by PFT. Each symbol is an individual plant. Lines are generalised additive model fits. (b) and (c) $M_F/M_S$ and leaf area ratio ($A_F/M_S$) at the average plant height in the dataset, estimated from fitted models in panel (a). Error bars are 95\% confidence intervals. Letters denote significant differences (at $\alpha = 0.05$).}
    \label{fig:figlmf}
\end{figure}
